---
title: "R Notebook"
output: html_notebook
---

```{install.packages(c("ggplot2", "dplyr", "shiny","reshape2", "tidyr", "car", "lmtest", "readr", "xts", "forecast", "tm", "knitr", "rmarkdown"))}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rmarkdown)
library(shiny)
library(readr)
```

---

---

R-exam project will focus on analyzing and understanding the connection between social economics factors like lunch type,test preparation, and students output or performance.

Research Question:

How do socio-economic factors, specifically parental education, lunch type, and completion of test preparation courses, influence students' academic performance in traditional exams, and what nuanced connections can be identified between socio-economic background and educational outcomes?

Project Abstract:

This project aims to explore the impact of socio-economic factors on students' academic performance, focusing on the StudentsPerformance dataset. Leveraging R programming, the investigation will focus on how variables such as parental education, lunch type, and completion of test preparation courses influence traditional exam scores.The research question delves into understanding the nuanced connections between socio-economic background and academic outcomes. Through statistical analyses and coding techniques in R, I aim to provide insights into educational disparities. The project adheres to a structured format, including an introduction, data description, methodology, results, and conclusion. By employing visualization tool aimed to present a comprehensive analysis.

```{# 1.Data exploration and cleaning using (tidyveres and dplyr)}
# Set the path to your CSV file
csv_file_path <- "StudentsPerformance.csv"

# Read the CSV file into a data frame
Student_data <- read.csv(csv_file_path)

# Print the first few rows of the data frame
head(Student_data)
# Exploring the dataset structure
str(Student_data)

# View summary statistics
summary(Student_data)

# Reading the first few rows of the dataset
head(Student_data)

missing_values <- Student_data %>%
  summarise_all(~ sum(is.na(.)))
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}

```

rrr

```{r}

```
