---
title: "The influence of  social-economic factors on  academic performance"
author: "Viola Awor"
date: "2023-12-17"
output: html_document
---


Data Libraries
```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rmarkdown)
library(shiny)
library(readr)
library(knitr)
library(reshape2)
```

1. Loading  student performance Dataset to read the first few rows and  exploring the structure  and datatype and cleaning the datset by checking for missing values if any. This Dataset doesnot have missing values
```{r}
csv_file_path <- "StudentsPerformance.csv"

# Read the CSV file into a data frame
Student_data <- read.csv(csv_file_path)

# Print the first few rows of the data frame
head(Student_data)
# Exploring the dataset structure
str(Student_data)

# View summary statistics and structure 
summary(Student_data)

# Reading the first few rows of the dataset to get understanding of the dataset
head(Student_data)

missing_values <- Student_data %>%
  summarise_all(~ sum(is.na(.)))
```
2.Statistical analysis to understand the relationships between socio-economic factors (parental education, lunch type, test preparation course completion) and students' performance.
1.Computed the  descriptive statistics for exam scores; Math, readingand writing.
Count: To know the total score for each exam.
Mean:To determine the average score for each exam and performance.
Standard deviation:To determine  of spread from the mean of the exam scores.
Minimum:To know the minimum score for exam
First quartile:To know the percentange composition of exam scores in quarters 25%
median:To know the score value separating the higher half from the lower half of each exam score.
Third quartile:To know the percentange composition of exam scores in quarters 75%
Maximum:To know the maximum score for each exam

```{r}
#2.Compute the discreptive statistics using dplr library

# Descriptive statistics for exam scores
# Using summarise function to compute calculate descriptive statistics: mean,Sd,min,count,median,Q1,Q3 and max values for  exam scores
describe_data <- function(column) {
  stats <- Student_data %>%
    summarise(
      Count = n(),
      Mean = mean(column, na.rm = TRUE),
      Standard_deviation = sd(column, na.rm = TRUE),
      Min = min(column, na.rm = TRUE),
      First_quartile_25th = quantile(column, 0.25, na.rm = TRUE),
      Median = median(column, na.rm = TRUE),
      Third_quartile_75th = quantile(column, 0.75, na.rm = TRUE),
      Max = max(column, na.rm = TRUE)
    )
  return(stats)
}

# Descriptive statistics for math score 
math_stats <- describe_data(Student_data$math.score)
print("Math Score Statistics:")
print(math_stats)

# Descriptive statistics for reading score
reading_stats <- describe_data(Student_data$reading.score)
print("Reading Score Statistics:")
print(reading_stats)

# Descriptive statistics for writing score
writing_stats <- describe_data(Student_data$writing.score)
print("Writing Score Statistics:")
print(writing_stats)




```
Computing the counts of  socio-economic categorical varibles to know the proportion of students according to each categorical variable

```{r}
# Computing counts for categorical  socio-economic variables

# Parental level of education
parental_education_counts <- table(Student_data$parental.level.of.education)

# Lunch type
lunch_type_counts <- table(Student_data$lunch)

# Test preparation course completion
test_prep_course_counts <- table(Student_data$`test.preparation.course`)

# outputting the results
print("Parental Level of Education Counts:")
print(parental_education_counts)

print("Lunch Type Counts:")
print(lunch_type_counts)

print("Test Preparation Course Completion Counts:")
print(test_prep_course_counts)
```

Plotting a boxplot graph to show comparison in exam score performance according to ethnicity

```{r}
#Boxplot showing comparison in performance between female and male students in exam scores.

# Reshape the data into a format  for easy plotting by ggplot2
long_data <- Student_data %>%
  gather(key = "subject", value = "score", math.score, reading.score, writing.score)

# Plotting the box plot graph
ggplot(long_data, aes(x = subject, y = score, fill = race.ethnicity)) +
  geom_boxplot() +
  facet_wrap(~ subject, scales = "free") +
  labs(title = "Performance Comparison by Ethnicity across different Subjects",
       x = "Subject",
       y = "Score") +
  theme_minimal()
```
Plotting a  grouped bar graph to show the comparison in average exam score performance according to gender

```{r}
#Bargraph showing average score by gender  across  subjects.
# Reshape the data for plotting
long_data <- Student_data %>%
  gather(key = "subject", value = "score", math.score, reading.score, writing.score)
# Plotting a grouped bar graph
ggplot(long_data, aes(x = subject, y = score, fill = gender)) +
  geom_bar(stat = "summary", fun.y = "mean", position = position_dodge()) +
  labs(title = "Average Scores by Gender across Subjects",
       x = "Subject",
       y = "Average Score") +
  theme_minimal()

```
A histogram  showing the distribution of the spread and skewness of the scores for all students.

```{r}
# Plotting histogram for Math Scores
ggplot(Student_data, aes(x = math.score)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Math Scores", x = "Math Score", y = "Count") +
  theme_minimal()
```
Histogram showing distribution of  reading score


```{r}
# Plotting histogram for Reading Scores
ggplot(Student_data, aes(x = reading.score)) +
  geom_histogram(bins = 30, fill = "red", alpha = 0.7) +
  labs(title = "Histogram of Reading Scores", x = "Reading Score", y = "Count") +
  theme_minimal()
```
Histogram showing distribution of writing score

```{r}
# Plotting histogram for Writing Scores
ggplot(Student_data, aes(x = writing.score)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  labs(title = "Histogram of Writing Scores", x = "Writing Score", y = "Count") +
  theme_minimal()
```
Pic chart representing proportion  distribution of students for each category of categorical variables.
```{r}
# Pie Chart showing proportion of students by Parental Level of Education with count figures
Student_data %>%
  count(parental.level.of.education) %>%
  ggplot(aes(x = "", y = n, fill = parental.level.of.education)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5)) +
  labs(title = "Proportion of Students by Parental Level of Education", fill = "Parental Education") +
  theme_void()
```
Pie Chart showing proportion distribution of students by Lunch Types
```{r}
# Pie Chart showing proportion of students by Lunch Types
Student_data %>%
  count(lunch) %>%
  ggplot(aes(x = "", y = n, fill = lunch)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5)) +
  labs(title = "Proportion of Students by Lunch Type", fill = "Lunch Type") +
  theme_void()
```
Pie Chart  showing proportion of students by Test Preparation Course Completion

```{r}
# Pie Chart  showing proportion of students by Test Preparation Course Completion
Student_data %>%
  count(`test.preparation.course`) %>%
  ggplot(aes(x = "", y = n, fill = `test.preparation.course`)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5)) +
  labs(title = "Proportion of Students by Test Preparation Course Status", fill = "Test Preparation Course") +
  theme_void()
```
Using dplyr library to
Showing the distribution of scores by  categorizing as (low, medium, high) across different categorical variables such as parental level of education, lunch type and test preparation course.

```{r}


#First, (categorize the scores into 'Low', 'Medium', and 'High')
#Then, plot stacked bar graphs for each subject score (Math, Reading, Writing) by parental level of education,test preparation course and lunch type

# Function to categorize scores <60,<80 and >=80 The categorize_score function assigns a category to each score.
categorize_score <- function(score) {
    if (score < 60) {
         "Low"
       } else if (score <= 80) {
           "Moderate"
         } else {
             "High"
           }
   }
# using  sapply  function to create new columns for categorized scores
Student_data$math_score <- sapply(Student_data$math.score, categorize_score)
Student_data$reading_score <- sapply(Student_data$reading.score, categorize_score)
Student_data$writing_score <- sapply(Student_data$writing.score, categorize_score)

```
Using the categorised distribution  to Plot a  stacked bar graph for Math Scores according to  Parental Level of Education.

```{r}
Student_data$math_score <- factor(Student_data$math_score, levels = c("Low", "Moderate","High"))
   ggplot(Student_data, aes(x = parental.level.of.education, fill=math_score)) +
     geom_bar(position = "fill") +
     labs(title = "Distribution of Math Scores by Parental Level of Education",
         x = "Parental Level of Education", y = "Proportion") +
     theme_minimal() +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Plotting a stacked bar graph for Writing scores according to lunch type
```{r}
  Student_data$writing_score <- factor(Student_data$writing_score, levels = c("Low", "Moderate","High"))
ggplot(Student_data, aes(x = lunch, fill=writing_score)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Writing scores by Lunch type",
       x = "Lunch type", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Plotting stacked bar graph for reading Scores by Parental Level of Education
```{r}
Student_data$reading_score <- factor(Student_data$reading_score, levels = c("Low", "Moderate","High"))
ggplot(Student_data, aes(x = test.preparation.course , fill=reading_score)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Reading Scores by Test preparation Course",
       x = "Test preparation Course", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Computing the  correlation coefficients between numeric variables( Exam scores)


```{r}
cor_matrix <- cor(Student_data[, c("math.score", "reading.score", "writing.score")], use = "complete.obs")
#library(reshape2)  to reshape the correlation matrix into a format suitable for ggplot to plot 
# Melt the data to suit the  ggplot format
melted_cor_matrix <- melt(cor_matrix)

```
Plotting a graph to visualize the correlation coeffcience of the numeric varibels.Showing a positive correlation between the exam scores starting from 0.5


```{r}
# heatmap graph showing Correlation Coefficients
ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Heatmap of Correlation Coefficients", x = "Variable 1", y = "Variable 2") +
  coord_fixed()
```
Testing  the hypotheses related to the impact of parental education, lunch type, and test preparation on exam score particularly math score by Formulating hypotheses and conducting t-tests and  ANOVA tests .
Evaluating p-values and making statistical inferences.

ANOVA for Parental Education Impact on Math Scores because it has more than two variables
```{r}
anova_result_parental_education <- aov(math.score ~ parental.level.of.education, data = Student_data)
summary(anova_result_parental_education)
```
# T-test for Lunch Type Impact on Math Scores (assuming two types of lunch)
```{r}
t_test_result_lunch_type <- t.test(math.score ~ lunch, data = Student_data)
summary(t_test_result_lunch_type)
```
T-test for Test Preparation Impact on Math Scores
```{r}
t_test_result_test_preparation <- t.test(math.score ~ test.preparation.course, data = Student_data)
summary(t_test_result_test_preparation)
```
Performing  regression analysis to model the relationship between socio-economic factors and academic performance by building regression models and assessing model fit.
 Linear regression model for Math Score
```{r}
#using lm function to create the regression model.
math_model <- lm(math.score ~ parental.level.of.education + lunch + test.preparation.course+ race.ethnicity, data = Student_data)
summary(math_model)

```
 Ceating linear regression model for Reading score using ml function
```{r}
reading_model <- lm(reading.score ~ parental.level.of.education + lunch + test.preparation.course+ race.ethnicity, data = Student_data)
summary(reading_model)
```
 Creating linear regression model for Writing score using lm function
```{r}
writing_model <- lm(writing.score ~ parental.level.of.education + lunch + test.preparation.course+ race.ethnicity, data = Student_data)
summary(writing_model)
```
 Generating predictions of exam scores for each model using predict function
```{r}
math_predictions <- predict(math_model, newdata = Student_data)
reading_predictions <- predict(reading_model, newdata = Student_data)
writing_predictions <- predict(writing_model, newdata = Student_data)

```
Reshaping the  dataframes  to format usable by ggplot to plot the predictions above.
```{r}
plot_data_math <- data.frame(Actual = Student_data$math.score, Predicted = math_predictions)
plot_data_reading <- data.frame(Actual = Student_data$reading.score, Predicted = reading_predictions)
plot_data_writing <- data.frame(Actual = Student_data$writing.score, Predicted = writing_predictions)

```
 Plotting a graph  to show Predicted vs. Actual values for Math Scores
```{r}
ggplot(plot_data_math, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs. Actual Math Scores", x = "Actual Scores", y = "Predicted Scores") +
  theme_minimal()
```

Plotting Predicted vs. Actual values for reading Scores
```{r}
ggplot(plot_data_reading, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs. Actual reading Scores", x = "Actual Scores", y = "Predicted Scores") +
  theme_minimal()
```
Plotting Predicted vs. Actual values for writing Scores
```{r}
ggplot(plot_data_writing, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs. Actual wrting Scores", x = "Actual Scores", y = "Predicted Scores") +
  theme_minimal()
```
Coefficient plot for Math Model
```{r}
coef_data <- broom::tidy(math_model)
ggplot(coef_data, aes(estimate, term)) +
  geom_point() +
  labs(title = "Coefficients of Math Score Model")
```

Coefficient plot for reading Model
```{r}
coef_data <- broom::tidy(reading_model)
ggplot(coef_data, aes(estimate, term)) +
  geom_point() +
  labs(title = "Coefficients of Reading Score Model")
```
Coefficient plot for writing Model
```{r}
#Using broom and tidy functions to transform thestatistical test into tidy datafram usable by ggplot
coef_data <- broom::tidy(writing_model)
ggplot(coef_data, aes(estimate, term)) +
  geom_point() +
  labs(title = "Coefficients of Writing Score Model")
````
Residual Plot for math model

```{r}
#Initialize the object
ggplot(math_model, aes(.fitted, .resid)) +# set mapping for the plot getting difference between the predicted values and residual value
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot for Math Score Model", x = "Fitted Values", y = "Residuals")
```
Residual Plot for Reading Model
```{r}
ggplot(reading_model, aes(.fitted, .resid)) + #set mapping for the plotgetting difference between the predicted values and residual value
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot for Reading Score Model", x = "Fitted Values", y = "Residuals")

```
Residual Plot for Writing Model
```{r}
ggplot(writing_model, aes(.fitted, .resid)) + # set mapping for the plot,getting difference between the predicted values and residual value
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot for Writing Score Model", x = "Fitted Values", y = "Residuals")
```
Comparing the impact of socio-economic factors on  exam scores.
Using side-by-side box plots or a paired t-test to compare scores.

Box plot  Impact of  Parental level of education on math score
```{r}
ggplot(Student_data, aes(x = parental.level.of.education, y = math.score, fill = parental.level.of.education)) +
  geom_boxplot() +
  labs(title = "Math Scores by Parental Education", x = "Parental Education", y = "Math Score") +
  theme_minimal()
```
Box plot showing Impact of  Parental level of education on Reading score
```{r}
ggplot(Student_data, aes(x = parental.level.of.education, y = reading.score, fill = parental.level.of.education)) +
  geom_boxplot() +
  labs(title = "Reading Scores by Parental Education", x = "Parental Education", y = "reading Score") +
  theme_minimal()
```
Box plot showing Impact of  Parental level of education on Reading score

```{r}
ggplot(Student_data, aes(x = parental.level.of.education, y = writing.score, fill = parental.level.of.education)) +
  geom_boxplot() +
  labs(title = "Writing Scores by Parental Education", x = "Parental Education", y = "writing Score") +
  theme_minimal()
```
 Box plot showing Impact of  lunch type on  math score
```{r}
# Box plot showing Impact of  lunch type on  math score
ggplot(Student_data, aes(x = lunch, y = math.score, fill = lunch)) +
  geom_boxplot() +
  labs(title = "Math Scores by Lunch Type", x = "Lunch Type", y = "Math Score") +
  theme_minimal()
```
Box plot showing Impact of  lunch type on  Reading score
```{r}
#Box plotBox plot showing Impact of  lunch type on  reading score
ggplot(Student_data, aes(x = lunch, y = reading.score, fill = lunch)) +
  geom_boxplot() +
  labs(title = "Reading Scores by Lunch Type", x = "Lunch Type", y = "reading Score") +
  theme_minimal()
```
Box plot showing Impact of  lunch type on  Reading score
```{r}
# Box plot Box plot showing Impact of  lunch type on  writing score
ggplot(Student_data, aes(x = lunch, y = writing.score, fill = lunch)) +
  geom_boxplot() +
  labs(title = "Writing Scores by Lunch Type", x = "Lunch Type", y = "writing Score") +
  theme_minimal()
```
Test preparation Course
Box plot showing the impact of test preparation course on math score
```{r}
ggplot(Student_data, aes(x = test.preparation.course, y = math.score, fill = test.preparation.course)) +
  geom_boxplot() +
  labs(title = "Math Scores by Test Preperation Course", x = "Test Preperation Course", y = "Math Score") +
  theme_minimal()
```
Box plot showing the impact of test preparation course on reading score
```{r}
ggplot(Student_data, aes(x = test.preparation.course, y = reading.score, fill = test.preparation.course)) +
  geom_boxplot() +
  labs(title = "Reading Scores by Test Preperation Course", x = "Test Preperation Course", y = "Reading Score") +
  theme_minimal()
```

Box plot showing the impact of test preparation course on writing score
```{r}

ggplot(Student_data, aes(x = test.preparation.course, y = writing.score, fill = test.preparation.course)) +
  geom_boxplot() +
  labs(title = "Writing Scores by Test Preperation Course", x = "Test Preperation Course", y = "Writing Score") +
  theme_minimal()
```
Dashboard to allow users easily interact with the data and insights


```{r}

# User interface Definition
ui <- fluidPage(
  titlePanel("Student Performance Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput("scoreType", "Select Score Type", 
                  choices = c("Math", "Reading", "Writing")),
      selectInput("socioFactor", "Select Socio-Economic Factor", 
                  choices = c("Parental Education", "Lunch Type", "Test Preparation","Ethnicity")),
      actionButton("submit", "Update Results")
    ),
    mainPanel(
      plotOutput("scorePlot"),
      tableOutput("dataTable")
    )
  )
)

# Server Logic
server <- function(input, output) {
  # Reactive expression to handle data based on user input
  filteredData <- reactive({
    # Loading the dataset
    Student_data <- read.csv("StudentsPerformance.csv")

    # Transform or filter data as needed
    Student_data <- data %>%
    rename(MathScore = math.score, ReadingScore = reading.score, WritingScore = writing.score)

  # 2. Creating a new variable, e.g., average score
  Student_data <- data %>%
    mutate(AverageScore = (math.score + reading.score + writing.score) / 3)

  # 3. Converting categorical variables to factors
  Student_data <- data %>%
    mutate(ParentalEducation = as.factor(ParentalEducation),
           LunchType = as.factor(LunchType),
           TestPreparation = as.factor(TestPreparation))

   return(Student_data) 
  })

  observeEvent(input$submit, {
    # Generate plot based on user input
    output$scorePlot <- renderPlot({
      Student_data <- filteredData()
      
      # Determine the y-axis variable based on input
      y_column <- switch(input$scoreType,
                         "Math" = "math.score",
                         "Reading" = "reading.score",
                         "Writing" = "writing.score")
      
      ggplot(Student_data, aes_string(x = input$socioFactor, y = y_column)) + 
        geom_point() +
        labs(title = paste(input$scoreType, "Scores by", input$socioFactor),
             x = input$socioFactor,
             y = paste(input$scoreType, "Score"))
    })

    # Output a table of the data
    output$dataTable <- renderTable({
      Student_data <- filteredData()
    })
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```
`
```
